Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-01-18 response-to-reviews.Rmd

*[issue #83](https://github.com/EcoClimLab/vertical-thermal-review/issues/83)*

I like your figures, but maybe consider homogenizing a little bit the styles for the plots between figures 2 3 and 4.

L205-206 Maybe add some references. How many degrees? In your Figure 3 there is not really a difference between Tleaf and Tair when SWR = 0. Given that the transpiration nearly stops at night (high humidity, low conductance) I would expect Tleaf – Tair to be very close to zero.

**Tleaf can be well below Tair on clear nights despite low transpiration, due to radiative coupling with the very cold sky. Added the following references to L205-206:**
**Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$)[@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020]**

@@

l258 under soil wet conditions?

*doing this because I don't think we want to talk about soil conditions. I've always had atmospheric conditions in my mind as I've thought about this.*

L298-299 Far red to red light: I think this could be developed here or somewhere else. You covered the vertical variation in light quantity inside the canopy, but you could also report the vertical change in light quality (light spectrum) as the leaves absorb preferentially some wavelengths but reflects or transmit others.

**Added the following lines:**
**Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers. Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000]**

@@

*[issue #95](https://github.com/EcoClimLab/vertical-thermal-review/issues/95)*

I was eager to see the part of the manuscript about photoprotection and also its relationship to heat damage. The opening sentence (l. 384) was obvious to the point of being painful to read.  Of course photoprotection is higher in the upper canopy: radiation levels are higher! Can this first sentence get revised to increase information content? This is one of the more exciting topics of the review, yet is handled only in a cursory fashion. If the authors could manage a bit more discussion of this topic it would be useful – instead they move (in the same paragraph) to VOC emissions, something I view as a separate capability of some plants. This is also confused in Fig. 1 where VAZ and VOC are placed on the same line. I’d like to see a more complete discussion of photoprotection and heat dissipation by NPQ first, and a separate paragraph about VOC. Also, there is nothing about critical temperature until late in the manuscript even though the authors skirt the issue in their text about photoprotection. This kind of thermal sensitivity is different from gas exchange (the section where the text about Tcrit occurs, l. 530-555), and should have been discussed earlier in the manuscript.

*[issue #92](https://github.com/EcoClimLab/vertical-thermal-review/issues/92)*
**It is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.**

@@

In the first part of Section 5.1, the authors conclude this paragraph stating the 'dominant role of vertical profiles in microclimate in shaping tree growth rates'. However, I believe the text has confused the very large effect of light with other aspects of forest microclimate, and they very much need to clarify this. If they believe other aspects of microclimate apart from light are drivers, then I believe this would be a minority opinion in the forest biology community. If they wish to speculate about other microclimate factors that cannot be disentangled from light, I suggest that they clearly indicate their opinion and speculation. It is perhaps an area for future research to disentangle light microclimate from aspects of vertical microclimate.

**We did not intend to argue that light was not the dominant factor, but the wording was misleading. This sentence has been revised to read, "This points to a dominant role of vertical profiles in the biophysical environment, particularly light (Fig. 2) in shaping tree growth rates within forests."**

@@

**Done**

line404-405: A cause-effect for an earlier, juvenile understory tree leaf-out is implied here, and I;m not sure that is demonstrated. There are other phenomena underlying this e.g. turgor, hydraulics, hormones, etc.

line475: 'because water supply often cannot meet the demands incurred …'. How do we know how frequent this is? Evidence?

**We have added four references showing that stomatal conductance is often limited by imbalance of water supply and evaporative demand for leaves in more elevated or more sunlit canopy positions [@yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010].**

@@

line478: 'intracellular CO_2' needs to be fixed.

**Done**

line479: 'In sun leaves, Tleaf thus further increases'. Further than what? Comparatives should have be clear what we’re comparing to.

**This sentence has been rewritten as, “Stomatal closure reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads.”**

@@

line481 'is maximized' or is maximal? Please correct.

**We have replaced “maximal” with “greatest”.**

line485-487: It seems that VPD and its effect on gs is ignored here. Please clarify? The well-known stomatal closure at high VPD alone would mean higher Tair would decrease gs in the upper canopy (not 'canopy leaves' as stated) compared to the lower canopy. See Grossiord et al. (2020) Tansley Review. VPD only gets mentioned in the section on photosynthesis (l. 496). Please do so earlier.

**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text underlined): “However, ~~because~~ water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al. 1994, Koch et al. 2004, Sillett et al 2010) and/or because of increased evaporative demand (VPD) caused by leaf warming in sunlit canopy locations (Buckley et al. 2014)***. As a result, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table 2), which drives the lower intracellular CO2 and carbon isotope discrimination discussed in section 3.1 (Table 1).  Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads ~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in gs in response to rising Tair”, and clarified that “high Tair should decrease gs of ***upper canopy***leaves”.**

### 2022-01-18 vertical-thermal-review.Rmd
added Tom's response to reviwer's comments and removed thermal capacitance in the sentence

Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`). 
Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. 
Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do lower-canopy leaves, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002]. 
The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020]. One  of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the canopy of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018]. However, the opposite trend was observed for a lone tree surrounded by grass in the same area, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients.

@@

Leaf hydraulic, stomatal and boundary layer conductances are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the upper canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
Maximum $g_s$ increases with light--typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979]--and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, water supply cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves with fully open stomata because, of height-related constraints on water transport [ **e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010**] and/or because of increased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations (Buckley et al. 2014). These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven by increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021]. This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.

>>>>

Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`). 
Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. 
Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do lower-canopy leaves, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002]. 
The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020]. One  of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the canopy of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018]. However, the opposite trend was observed for a lone tree surrounded by grass in the same area, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients.

@@

Leaf hydraulic, stomatal and boundary layer conductances are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the upper canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
Maximum $g_s$ increases with light--typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979]--and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, water supply cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves with fully open stomata because, of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or because of increased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations (Buckley et al. 2014). These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven by increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021]. This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.
